---
title: "Velocity in 3D and transition probability analysis"
author: "Joseph Ng"
date: "26/06/2020"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(plyr)
library(rgl)
library(Seurat)
library(SeuratWrappers)
```

This markdown document contains codes used to plot `velocyto` results on 3D UMAP projections, and quantification of transition probability between clusters.

In the beginning we also attach the code used to remove Ig, TCR related genes from the variably-expressed features to avoid these genes to bias dimensionality reduction - this might be a useful line for those working with similar cell types.

```{r loadData, echo=FALSE}
pure <- readRDS('../Pure2_finalvs_UMAP3D.rds')

# remove Ig, TCR related genes from variable features
VariableFeatures(pure) <- VariableFeatures(pure)[!grepl("^IGH[MDE]|^IGHG[1-4]|^IGHA[1-2]|^IG[HKL][VDJ]|^IGKC|^IGLC[1-7]|^TR[ABGD][CV]|^AC233755.1|^IGLL|^JCHAIN", VariableFeatures(pure))]
pure@active.assay <- 'RNA'
pure

# write cluster identities as a column in metadata
pure[['cluster']] <- Idents(pure)

GetClusterCentroids <- function(emb, idents)
{
  # function to fetch coordinates of cluster centroids from dimensionality-reduction space
  emb <- split(as.data.frame(emb), f = idents)
  o <- do.call("rbind", lapply(emb, function(x) apply(x, MARGIN = 2, median, na.rm = TRUE)))
  o <- as.data.frame(o)
  o$celltypes <- names(emb)
  o
}
```

```{r velocity_plotFx, include=FALSE}
PrepareCellsForVelocity <- function(object, idents, reduction = "umap", dims = c(1, 2),
                                    clean = TRUE, clean_perc = 0.99)
{
  # Select cells of given identities for plotting velocity.
  # Optinally clean cells by removing cells too far away from the rest of 
  # the cells of the same cluster. Default use 99% (clean.perc = 0.99) percentile -
  # cells with distance from cluster centroid beyond this percentile are removed.
  # This is to avoid cells which are outliers of a given cluster to bias the overall 
  # velocity depiction (otherwise they might lead to artefact velocity streams different
  # from the general flow)
  cell.vec <- Cells(object)[which(Idents(object) %in% idents)]
  ident.vec <- Idents(object)[which(Idents(object) %in% idents)]
  emb <- Seurat::Embeddings(object, reduction = reduction)[cell.vec, ]
  if(clean){
    centroids <- GetClusterCentroids(emb, ident.vec)
    dist_from_centroids <- do.call("rbind", lapply(cell.vec, function(cell){
      id <- as.character(ident.vec[cell])
      cell.coord <- unlist(emb[cell, ])
      centroid.coord <- unlist(centroids[which(centroids$celltype == id), 1:ncol(emb)])
      eucl <- ( (cell.coord[1] - centroid.coord[1])^2 + 
                  (cell.coord[2] - centroid.coord[2])^2 +
                  (cell.coord[3] - centroid.coord[3])^2 ) ^ 0.5 # euclidean distance
      data.frame(CB = cell, celltype = id, distance = eucl, stringsAsFactors = FALSE)
    }))
    allowed_dist_from_centroids <- do.call("rbind", lapply(idents, function(i){
      o <- dist_from_centroids[which(dist_from_centroids$celltype == i), "distance"]
      data.frame(celltype = i, limit = quantile(o, probs = clean_perc, na.rm = TRUE),
                 stringsAsFactors = FALSE)
    }))
    dist_from_centroids$allowed <- apply(dist_from_centroids, MARGIN = 1, function(x){
      limit <- allowed_dist_from_centroids[which(allowed_dist_from_centroids$celltype == x[2]),
                                           "limit"]
      return(as.numeric(x[3]) < limit)
    })
    allowed_cells <- dist_from_centroids[which(dist_from_centroids$allowed), "CB"]
    emb <- emb[allowed_cells, ]
  }
  emb[, dims]
}

```

## Processing velocity results

We follow [this](https://github.com/satijalab/seurat-wrappers/blob/master/docs/velocity.md) vignette from `SeuratWrappers`. 

Below, `vel` is the object `Tool(object = bm, slot = "RunVelocity")` from the velocyto calculation demonstrated in the vignette above.

```{r velocity_calc, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=4}
load('../scPure_withIgE_velocity.RData')
colours <- Cells(pure)
palette <- scales::hue_pal()(10)[c(5,2,9,1,6,4,10,8,7,3)]#[c(4, 6, 10, 3, 7, 8, 1, 9, 2, 5)]
names(palette) <- levels(Idents(pure))
for(celltype in levels(Idents(pure))){
  colours <- replace(colours, which(Idents(pure) == celltype), palette[celltype])
}
names(colours) <- Cells(pure)
# plot here
par(mfrow = c(1, 3))
em <- PrepareCellsForVelocity(pure, idents = levels(Idents(pure)),
                              reduction = "umap", dims = c(1, 3), 
                              clean = TRUE, clean_perc = 0.9)
vel_results_13 <- velocyto.R::show.velocity.on.embedding.cor(emb = em, vel = vel, 
                                                             n = 200, scale = "sqrt", 
   cell.colors = velocyto.R::ac(x = colours[rownames(em)], alpha = 0.7), 
   cex = 0.8, arrow.scale = 1.5, show.grid.flow = TRUE, min.grid.cell.mass = 10, 
   grid.n = 40, arrow.lwd = 1, do.par = FALSE, cell.border.alpha = 0, return.details = TRUE,
   xlab = "UMAP1", ylab  = "UMAP3")

em <- PrepareCellsForVelocity(pure, idents = levels(Idents(pure)),
                              reduction = "umap", dims = c(2, 3), 
                              clean = TRUE, clean_perc = 0.9)
vel_results_23 <- velocyto.R::show.velocity.on.embedding.cor(emb = em, vel = vel, 
                                                             n = 200, scale = "sqrt", 
   cell.colors = velocyto.R::ac(x = colours[rownames(em)], alpha = 0.7), 
   cex = 0.8, arrow.scale = 1.5, show.grid.flow = TRUE, min.grid.cell.mass = 10, 
   grid.n = 40, arrow.lwd = 1, do.par = FALSE, cell.border.alpha = 0, return.details = TRUE, 
   xlab = "UMAP2", ylab  = "UMAP3")

em <- PrepareCellsForVelocity(pure, idents = levels(Idents(pure)),
                              reduction = "umap", dims = c(1, 2), 
                              clean = TRUE, clean_perc = 0.9)
vel_results_12 <- velocyto.R::show.velocity.on.embedding.cor(emb = em, vel = vel, 
                                                             n = 200, scale = "sqrt", 
   cell.colors = velocyto.R::ac(x = colours[rownames(em)], alpha = 0.7), 
   cex = 0.8, arrow.scale = 1.5, show.grid.flow = TRUE, min.grid.cell.mass = 10, 
   grid.n = 40, arrow.lwd = 1, do.par = FALSE, cell.border.alpha = 0, return.details = TRUE, 
   xlab = "UMAP1", ylab  = "UMAP2")
```

# A method to quantify velocity stream to analyse (pairwise) transition between clusters

Here is the implementation of the method to quantify the strength of individual velocity streams (i.e. arrows on the velocity projections) into a transition score between pairs of clusters. This is done by the `count_arrows` function below.

A plotting function (`plotTransition`) is provided to plot the transition scores.

```{r ,fig.width =12, fig.height=3}
Calc3DEuclDist <- function(pt1, pt2){
  sqrt( (pt1[1] - pt2[1]) ^ 2 + (pt1[2] -  pt2[2]) ^ 2 + (pt1[3] -  pt2[3]) ^ 2 )
}
count_arrows <- function(vel_results, SeuratObj, dims = c(1, 2)){
  centroids <- GetClusterCentroids(Embeddings(SeuratObj, reduction = "umap")[, dims], 
                                   idents = Idents(SeuratObj))
  centroids3d <- GetClusterCentroids(Embeddings(SeuratObj, reduction = "umap")[, 1:3], 
                                     idents = Idents(SeuratObj))
  rx <- range(c(range(as.data.frame(vel_results$arrows)$x0),
                range(as.data.frame(vel_results$arrows)$x1)))
  ry <- range(c(range(as.data.frame(vel_results$arrows)$y0),
                range(as.data.frame(vel_results$arrows)$y1)))
  stepsize_x <- seq(rx[1], rx[2], length.out = 50)
  stepsize_x <- stepsize_x[2] - stepsize_x[1]
  stepsize_y <- seq(ry[1], ry[2], length.out = 50)
  stepsize_y <- stepsize_y[2] - stepsize_y[1]

  arrow_test <- do.call("rbind", apply(vel_results$garrows, MARGIN = 1, function(y){
    nearby_cells <- as.data.frame(Embeddings(SeuratObj, reduction = "umap")[, dims])
    nearby_cells <- nearby_cells[nearby_cells[, 1] > (y[1] - stepsize_x / 2) &
                                 nearby_cells[, 1] < (y[1] + stepsize_x / 2) &
                                 nearby_cells[, 2] > (y[2] - stepsize_y / 2) &
                                 nearby_cells[, 2] < (y[2] + stepsize_y / 2) , ]
    nearby_cells <- Idents(SeuratObj)[rownames(nearby_cells)]
    majority <- table(nearby_cells)
    k <- names(which.max(majority))
    o <- apply(centroids, MARGIN = 1, function(x){
      X <- y[1:2]
      Y <- y[3:4]
      Z <- as.numeric(unlist(x[1:2]))
      dist_1 <- sqrt( (X[1] - Y[1]) ^ 2 + (X[2] -  Y[2]) ^ 2 )
      dist_2 <- sqrt( (X[1] - Z[1]) ^ 2 + (X[2] -  Z[2]) ^ 2 )
      dist_3 <- sqrt( (Y[1] - Z[1]) ^ 2 + (Y[2] -  Z[2]) ^ 2 )
      sc <- 1 - acos( (dist_1^2 + dist_2^2 - dist_3^2) / (2 * dist_1 * dist_2) )/ pi
      dist3d_btwCentroids <- Calc3DEuclDist(as.numeric(centroids3d[centroids3d$celltypes == unlist(x[3]),
                                                                   1:3]),
                                            as.numeric(centroids3d[centroids3d$celltypes == k,
                                                                   1:3]))
      sc * dist_1 / dist_2
#      if(dist3d_btwCentroids == 0) return(0)
#      else return(sc * dist_1 / ( dist_2 * dist3d_btwCentroids))
    })
    suppressWarnings(o$celltype <- k)
    as.data.frame(o)
  }))
  rownames(arrow_test) <- sapply(1:nrow(arrow_test), function(x) paste("arrow", x, sep = ""))
  arrow_test <- arrow_test[order(arrow_test[, ncol(arrow_test)]), ]
  annotation <- data.frame(arrow_test[, ncol(arrow_test)])
  rownames(annotation) <- rownames(arrow_test)
  arrow_test <- arrow_test[, 1:(ncol(arrow_test) - 1)]

  arrow_test_summarised <- data.frame(from = as.character(annotation[, 1]), 
                                      to = apply(arrow_test, 1, 
                                                 function(x) names(which.max(x))),
                                      score = apply(arrow_test, 1, max),
                                      stringsAsFactors = FALSE)
  arrow_test_summarised[, 2] <- gsub(".", "-", arrow_test_summarised[, 2], fixed= TRUE)
  arrow_test_summarised <- arrow_test_summarised[which(arrow_test_summarised[, 1] != 
                                                       arrow_test_summarised[, 2]), ]
  arrow_test_summarised
}

# calculate for every pairwise combination of UMAP dimensions
arrows_12 <- count_arrows(vel_results_12, SeuratObj = pure, dims = c(1, 2))
arrows_13 <- count_arrows(vel_results_13, SeuratObj = pure, dims = c(1, 3))
arrows_23 <- count_arrows(vel_results_23, SeuratObj = pure, dims = c(2, 3))
arrows_12$dim <- "1_2"
arrows_13$dim <- "1_3"
arrows_23$dim <- "2_3"

centroids3d <- GetClusterCentroids(Embeddings(pure, reduction = "umap"), Idents(pure))

# evaluate pairwise distances between clusters using the cluster centroids coordinates in 3D
pairwise_distance <- matrix(NA, nrow = length(levels(Idents(pure))),
                            ncol = length(levels(Idents(pure))),
                            dimnames = list(levels(Idents(pure)),
                                            levels(Idents(pure))))
for(i in 1:nrow(pairwise_distance)){
  for(j in 1:nrow(pairwise_distance)){
    pairwise_distance[i, j] <- Calc3DEuclDist(as.numeric(centroids3d[i, 1:3]),
                                              as.numeric(centroids3d[j, 1:3]))
  }
}
# normalise into the range [0,1]
pairwise_distance <- apply(pairwise_distance, MARGIN = 2, function(z){
  range_z <- range(z[z > 0])
  z <- replace(z, which(z == 0), NA)
  1 - (z - range_z[1]) / diff(range_z)
})


plotTransition <- function(tb, pwdist){
  # function to process the transition score and plot as grid
  arrow_count_summary <- ddply(tb, #arrows_12, 
                             .variables = c("from", "to"), summarise, 
                             total_score = sum(score))
  arrow_count_summary$total_score <- apply(arrow_count_summary, MARGIN = 1, function(x){
    as.numeric(x[3]) * pwdist[x[1], x[2]]
  })
  arrow_count_toPlot <- arrow_count_summary
  arrow_count_toPlot$total_score <- log2(1 + arrow_count_toPlot$total_score)
  ggplot(arrow_count_toPlot, aes(x = from, y = to)) + 
    geom_tile(aes(fill = total_score)) + theme_classic() + 
    scale_fill_gradient2() +
    #scale_fill_gradient2(low = "white", mid = "lavenderblush2", 
    #                     high = "purple3",  midpoint = 0.7) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), validate = TRUE)
}

# how does using different combinations of axes change the picture?
g1 <- plotTransition(arrows_13, pairwise_distance) + ggtitle("UMAP1v3")
g2 <- plotTransition(rbind(arrows_13, arrows_23), pairwise_distance) + ggtitle("UMAP1v3, 2v3")
g3 <- plotTransition(rbind(arrows_13, arrows_23, arrows_12), pairwise_distance) + 
  ggtitle("All pairwise UMAP axes")
gridExtra::grid.arrange(g1, g2, g3, ncol = 3)
```
